Parameter estimation

### 2022-08-14 SFE-report/Paramter_fitting.tex
14.08

As discussed by \citet{Himmelblau1970}, The Bayesian approach to estimation make use of prior information. Such prior knowledge can come from theoretical considerations, from the results of previous experiments, or from assumptions by the experimenter. Typically, a Bayesian approach assumes a prior probability distribution of an unknown parameter $\theta$ in some parameter space $\boldsymbol{\theta}$. The distribution is updated by using Bayes' rule to obtain the posterior probability distribution.

@@

Further, if al events comprising the set $\{A_i\}$ are included in $A_1,A_2,...,A_n$, then

@@

	{\footnotesize
	\begin{equation}
		P\left( A_1, \text{or~} A_2,..., \text{or~} A_n \right) = \sum_{i=1}^{n} P\left( A_i \right)
	\end{equation} }

@@

\newpage
\section{Parameter estimation} \label{CH: Parameter_estimation}

Conceptually, the unobservable $\epsilon(t)$ is added to $y(t)$ to give the observable dependent variable $Y(t)$. For discrete observations:

@@

If the estimated parameter $\hat{\theta}$ replaces the model parameter $\theta$, the residue error is $\textbf{E} = Y(t) - \hat{Y}(t)$, where $\hat{Y}(t)$ is the predicted value of $Y_i$ at $x_i$. The objective in parameter estimation is to obtain the "best" estimate of $\theta$ based on the continuous observations $Y(t)$ or the discrete observations $Y(t_i)$.

We assume that $x(t)$ is given \textit{a priori} but the $\textbf{y}_0$ and $\theta$ are to be estimated over the time interval $0 \leq t \leq t_n$ from discrete observation described by the following relation:

{\footnotesize
\begin{equation}
	\textbf{Y}(t_i) = \textbf{h}(t_i)y(t_i) + \boldsymbol{\epsilon}(t_i) \qquad 1 \leq i \leq n
\end{equation} }
where $\textbf{Y}(t_i)$ is an $n \times 1$ column vector, $\textbf{h}(t_i)$ is an $n \times v$ matrix given \textit{a priori}, and $ \boldsymbol{\epsilon} (t_i)$ is an $n \times 1$ column vector ("noise" vector) whose elements are the unobservable errors. The above equation can be written in general form as:

{\footnotesize
\begin{equation} \label{EQ: Measurment_noise}
	\textbf{Y}(t_i) = \Psi(\theta, y_0, t_i) + \epsilon(t_i)
\end{equation}}
where $\Psi$ is the matrix of functions.

Because of the difficulty of obtaining analytical solutions to the deterministic process model, experiments have been arranged wherby the vector of derivatives $d\textbf{Y}/dy$ is measured rather than \textbf{Y} itself. In such cases, it is assumed that the unobservable error is added to the deterministic derivative $dy/dt$ as follows:

{\footnotesize
\begin{equation}
	\cfrac{d\textbf{Y}}{dt} = \cfrac{dy}{dt} + \epsilon
\end{equation} }

\subsection{Least squares estimation}
A least squares parameter estimation does not require prior knowledge of the distribution of unobservable errors, yield  unbiased estimates, that is, $ \mathbb{E}\{\hat{Y}(t) = y(t) \} $, and results in the minimum variance among all linear unbiased estimators. If the observations $\textbf{Y}$ for the model responses are continuous functions of time from $t = 0$ to $t = t_f$, the Markov (or "rigorous least squares") criterion is to minimize:

{\footnotesize
\begin{equation}
	\phi =  \cfrac{1}{2} \int_{0}^{t_f}  \left[ \textbf{Y} - \Psi \right]^T \Gamma^{-1} \left[ \textbf{Y} - \Psi \right] dt
\end{equation} }
where $\Gamma$ is the covariance matrix (or perhaps a matrix of appropriate weights), and $\phi$ is the time integrated value of the error squared ("integral squared error"). If the observations are made at discrete instants of time, $t_i$, $i = 1,2,...,n$, the Markov criterion is to minimize:

{\footnotesize
\begin{equation}
	\phi = \cfrac{1}{2} \sum_{i=1}^{n} \left[ \textbf{Y}(t_i) - \Psi(t_i) \right]^T \Gamma^{-1} \left[ \textbf{Y}(t_i) - \Psi(t_i) \right] 
\end{equation} }

If $\Gamma$ is a diagonal matrix, $\phi$  becomes a "weighted least squares" criterion; if $\Gamma = \sigma^2 \textbf{I}$ where $\sigma$ is a standard deviation, $\phi$ is the "ordinary least squares" criterion.

@@

{\footnotesize
\begin{flalign*}
	&{\normalsize \text{Model A - independent error:} } &&\nonumber \\
	Y_i &= \Psi(\theta, y_0, t_i) + \epsilon_i \qquad i = 1,2,...,n&& \\
	&{\normalsize \text{Model B - cumulative data:}} && \nonumber \\
	Y_j - Y_{j-1} &= \Psi(\theta, y_0, t_j) - \Psi(\theta, y_0, t_{j-1}) + \epsilon_j, \quad j = 1,2,...,n&& \\
	&{\normalsize \text{or}} && \nonumber \\
	Y_i = \sum_{j=1}^{i} &\left( Y_j - Y_{j-1} \right) = \sum_{j=1}^{i} \left( \Psi(\theta, y_0, t_j) - \Psi(\theta, y_0, t_{j-1}) \right) + \sum_{j=1}^{i} \epsilon_j &&
\end{flalign*} }

in which $\epsilon_i$ are independent random variables with $\mathbb{E}\{\epsilon_i\}$ and Var$\{\epsilon_i\} = \sigma_1^2$, a constant. The $\epsilon_j$ are also random independent variables because they represent differences such as $\epsilon_3' - \epsilon_2' = \left( \epsilon_1 + \epsilon_2 + \epsilon_3 \right) - \left( \epsilon_1 + \epsilon_2 \right) = \epsilon_3$. We assume that $\mathbb{E}\{\epsilon_j\}$ and Var$\{\epsilon_i\} = \sigma_1^2\left(x_j - x_{j-1}\right)$; i.e., the variance of $\epsilon_j$ can be a function of the test interval. The experimental point tend to stay on one side of the regression line of the best fit for Model B.

The best unbiased estimate of the Model B is obtained by minimizing the weighted sum of squares:

{\footnotesize
\begin{equation*}
	\sum_{j=1}^{n} \left( \cfrac{1}{x_j - x_{j-1}} \right) \left[ \left( Y_j - Y_{j-1} \right) - \left( \Psi(\theta, y_0, t_j) - \Psi(\theta, y_0, t_{j-1}) \right) \right]^2 
\end{equation*} }

\subsection{Maximum likelihood estimation}
A well-know and desirable procedure (when it can be carried out) is that of maximum likelihood which leads asymptotically to estimates with the greatest efficiency but not necessarily unbiased. A desirable feature of the maximum likelihood methods is that, under certain conditions (which are not ntoo rigorous), the estimated parameters are normally distributed for large samples.

Suppose that $p(x; \theta_1, \theta_2,...,)$ is a probability density function of \textit{known} form for the random variable $X$, a function which contains one or more parameters $\theta_1, \theta_2,...$. Also, suppose that we do not know what the values of $\theta_1, \theta_2, ...$ are.How can the most suitable values of $\theta_1, \theta_2, ...$ be estimated? One way is to collect a random sample of values of the random variable $X$, $\{x_1,x_2,...x_n\}$ and select the values of $\theta_1, \theta_2, ...$ now regarded random variables, that maximize the likelihood function $L\left( \theta_1, \theta_2, ... | x_1,x_2,...,x_n \right)$, a function described in chapter \ref{CH: Bayes} in connection with Bayes' theorem. Such estimators, $\hat{\theta}_1,\hat{\theta}_2, ...$ are known as maximum likelihood estimators. In effect, the methods selects those values of $\theta_1, \theta_2, ...$ that are at least as likely to generate the observed sample as any other set of values of the parameters if the probability density of values of the parameters if the probability density of the random variable $X$ were to be extensively simulated through use if the probability density $p\left( x|\theta_1, \theta_2, ... \right)$. In making a maximum likelihood estimate, we assume that the form of the probability density is connect (only the $\theta$'s need be determined) and that all possible values of $\theta_i$ are equally likely before experimentation.

The likelihood function for the parameters given one observation is just the probability density in which the observation is regarded as a fixed number and the parameters as the variables:

{\footnotesize
\begin{equation}
	L\left( \theta_1, \theta_2,...|x_1 \right) = p\left( x_1; \theta_1, \theta_2, ... \right)
\end{equation} }
where the lower case $x$'s and the number subscripts designate the value of the respective observation that is inserted into the probability density function. The likelihood function for the parameters based on several observations is the product of the individual functions if the observations are independent:

{\footnotesize 
\begin{multline}
	L\left( \theta_1, \theta_2,...|x_1, x_2, ..., x_n \right) = \prod_{i=1}^{n} L\left( \theta_1, \theta_2,...|x_1 \right) \\
	= p\left( x_1; \theta_1, \theta_2, ... \right)p\left( x_2; \theta_1, \theta_2, ... \right)\cdots p\left( x_n; \theta_1, \theta_2, ... \right)
\end{multline} }

In choosing as estimates of $\theta_i$ the values that maximize $L$ for the given values $\left( x_1, x_2, ..., x_n \right)$, it turns out that it is more convenient to work with the $\ln L$ than with $L$ itself:

{\footnotesize 
\begin{align}
	\ln L &= \ln p \left( x_1; \theta_1, \theta_2, ... \right) + \ln p\left( x_3; \theta_1, \theta_2, ... \right) + ... \nonumber \\
	&= \sum_{i=1}^{n} \ln p\left( x_i; \theta_1, \theta_2, ... \right)
\end{align} }

The $\ln L$ can be maximized with respect to the vector $\theta$ by equating to zero the partial derivatives of $\ln L$ with respect to each of the parameters:

{\footnotesize
\begin{align}\label{EQ: MLE_Basic_Solution}
	\cfrac{\partial \ln L}{\partial \theta_1} &= \cfrac{\partial \sum_{i=1}^{n} \ln p\left( x_1; \theta_1, \theta_2, ... \right) }{\partial \theta_1} = 0 \nonumber \\
	\cfrac{\partial \ln L}{\partial \theta_2} &= \cfrac{\partial \sum_{i=1}^{n} \ln p\left( x_1; \theta_1, \theta_2, ... \right) }{\partial \theta_2} = 0
\end{align} }

Solution of Equations \ref{EQ: MLE_Basic_Solution} yield the desired estimates $\hat{\theta}_1, \hat{\theta}_2, ...$. (Often iterative rather than analytical methods must be employed to obtain vector \boldsymbol{$\hat{\theta}$}). By carrying out this operation, under fairly unrestrictive conditions, it can be shown that as $n$ approaches infinity the maximum likelihood estimates have the desirable asymptotic properties:

\begin{enumerate}[label=(\arabic*)]
	\item {\footnotesize $\lim\limits_{n \rightarrow \infty} \mathbb{E}\{\hat{\theta_i}\} = \theta_i$ }
	\item {\footnotesize $\left[ \sqrt{n}\left( \hat{\theta_i} - \theta_i \right) \right]$ } is normally distributed
\end{enumerate}
and for the case of two parameters:

\begin{enumerate}[resume,label=(\arabic*)]
	\item {\footnotesize $\lim\limits_{n \rightarrow \infty} \text{Var}\{\hat{\theta_i}\} = \cfrac{1}{n} \left[ \mathbb{E} \left\{ \left( \cfrac{\partial \ln p}{\partial \theta_i} \right)^2 \right\} \right]^{-1} \cfrac{1}{1-\rho^2_{\hat{\theta_1}, \hat{\theta_2}}} $ } \label{Assumption_MLE}
\end{enumerate}

where $\rho^2_{\hat{\theta_1}, \hat{\theta_2}}$ is the coefficient of correlation of the two $\hat{\theta}$'s. Extension of \ref{Assumption_MLE} to more than two parameters requires the use od matrix notation and will not be shown here.

Maximum likelihood estimates are not necessarily unbiased. Maximum likelihood estimates, however are efficient and, hence, consistent estimates. Furthermore, where a sufficient estimate can be obtained, the maximum likelihood method will obtain it. Finally, if $\hat{\theta}$ is a maximum likelihood estimator of $\theta$, the $f\left( \theta \right)$ is a maximum likelihood estimator of $f\left( \theta \right)$, a function of $\theta$.

@@

By assuming that the conditional distribution of $\bar{Y}_i$, given $x_i$, is normal about $\eta_i = \mathbb{E}\left\{ \bar{Y}_i | x_i \right\}$, then we form the likelihood function based on the probability density:

{\footnotesize
\begin{align} \label{EQ: MLE_Norm}
	p\left( y | x; \theta, \sigma^2_{\bar{Y}_i} \right) &= \cfrac{1}{\sqrt{2 \pi} \sigma^2_{\bar{Y}_i} } \exp \left[ \cfrac{1}{2  \sigma_{\bar{Y}_i} } \left( Y - \Psi(\theta_i) \right)^2 p_i \right] \nonumber \\
	L\left( \theta, \sigma_{\bar{Y}_i} | y,x \right) &\equiv L = \prod_{i=1}^{n} \cfrac{1}{\sqrt{2 \pi} \sigma^2_{\bar{Y}_i} } \exp \left[ \cfrac{1}{2  \sigma_{\bar{Y}_i} } \left( Y - \Psi(\theta_i) \right)^2 p_i \right]
\end{align}}

In Equation \ref{EQ: MLE_Norm} the parameters are the variables, and the values of $Y$ and $x$ are given. Then

{\footnotesize
	\begin{equation}
		\ln L = -n \ln \sqrt{2 \pi} - \cfrac{n}{2} \ln \sigma_{\bar{Y}_i}
		- \cfrac{ \sum_{i=1}^{n} p \left\{ \left[  \bar{Y}_i - \Psi(\theta_i) \right]^2 \right\} }{2 \sigma^2_{\bar{Y}_i}}
	\end{equation}
	}

To obtain the maximum likelihood estimates, we place

{\footnotesize
	\begin{equation}
		\cfrac{\partial \ln L}{\partial \theta} = \cfrac{\partial \ln L}{\partial \left( \sigma_{\bar{Y}^2_i} \right)} = 0
	\end{equation}
}

or iteratively solve

{\footnotesize
	\begin{equation}
		\hat{\theta}_{MLE} = \arg \max_{\theta \in \Theta} L = \arg \max_{\theta \in \Theta} p(\theta|y)
	\end{equation}
}

Maximum likelihood estimation (MLE) is a method of estimating the parameters of an assumed probability distribution, given some observed data. This is achieved by maximizing a likelihood function so that, under the assumed statistical model, the observed data is most probable. The MLE has the desirable characteristics of asymptotic efficiency and normality. Each time it has been associated with the (joint) normal distribution because of mathematical convenience. Consider the joint probability density function (the likelihood function) $p(\theta, y_0|y)$ for random sample $y = y(t_1),y(t_2),...,y(t_n)$, parameter $\theta$ and initial condition $y_0$ if a maximum of this function over all choices of $y_0$ and $\theta$ can be found, the estimates so obtained are maximum likelihood estimates. The conditions at the maximum can be evolved incorporating prior information as follows.

The posterior probability density $p(\theta,y_0|y)$ can be expressed as the ratio of two probability densities if we make use of the analogue for continuous variables of Equation \ref{EQ: Probabiltiy_independet_b}:

{\footnotesize
\begin{align} \label{EQ: Posterior_distribution}
	p\left( \theta,y_0|y \right) = \cfrac{p\left( \theta, y_0, y \right)}{p\left( y \right)}
\end{align} }

{\footnotesize
\begin{equation}
	\begin{split}
		p\left( \theta,y_0,y \right) &= p\left( y(t_n)|\theta,y_0,y(t_{n-1}),...,y(t_1) \right) \\ &\cdot p\left( \theta,y_0,y(t_{n-1}),...,y(t_1) \right)
	\end{split}
\end{equation} }

{\footnotesize
\begin{equation}
	p\left( \theta,y_0,y \right) = p\left( \theta,y_0 \right) \prod_{i=1}^{n} p\left( \theta,y_0|y \right)
\end{equation} }

Examination of Equation \ref{EQ: Measurment_noise} shows that $\textbf{Y}(t_i)$ depends only on $t_i$, $y_0$, $\theta$ and $\epsilon(t_i)$ and is not conditioned by any previous measurement. Consequently, we can write

>>>>

As discussed by \citet{Himmelblau1970}, the Bayesian approach to estimation make use of prior information. Such prior knowledge can come from theoretical considerations, from the results of previous experiments, or from assumptions by the experimenter. Typically, a Bayesian approach assumes a prior probability distribution of an unknown parameter $\theta$ in some parameter space $\boldsymbol{\theta}$. The distribution is updated by using Bayes' rule to obtain the posterior probability distribution.

@@

Further, if all events comprising the set $\{A_i\}$ are included in $A_1,A_2,...,A_n$, then

@@

	{\footnotesize
	\begin{equation}
		P\left( A_1 \text{or~} A_2... \text{or~} A_n \right) = \sum_{i=1}^{n} P\left( A_i \right)
	\end{equation} }

@@

\clearpage

\section{Parameter estimation} \label{CH: Parameter_estimation}

Conceptually, the unobservable error $\epsilon(t)$ is added to the deterministic model output, $y(t)$, to give the observable dependent variable $Y(t)$, for example the experimental process output. For discrete observations:

@@

%If the estimated parameter $\hat{\theta}$ replaces the model parameter $\theta$, the residue error is $\textbf{E} = Y(t) - \hat{Y}(t)$, where $\hat{Y}(t)$ is the predicted value of $Y_i$ at $y_i$. 
The objective in parameter estimation is to obtain the "best" estimate of $\theta$ based on the continuous observations $Y(t)$ or the discrete observations $Y(t_i)$. 
\sout{
We assume that $x(t)$ is given \textit{a priori} (for example as a solution of process model) but $\theta$ is to be estimated over the time interval $0 \leq t \leq t_n$ from discrete observation described by the following relation:}

{\footnotesize
\begin{equation}
	 Y(t_i) = \textbf{h}(t_i)y(t_i) + \boldsymbol{\epsilon}(t_i) \qquad 1 \leq i \leq n
\end{equation} }
\sout{where $ Y(t_i)$ is an $n \times 1$ column vector, $\textbf{h}(t_i)$ is an $n \times v$ matrix given \textit{a priori}, and $ \boldsymbol{\epsilon} (t_i)$ is an $n \times 1$ column vector ("noise" vector) whose elements are the unobservable errors. The above equation can be written in general form as:}

{\footnotesize
\begin{equation} \label{EQ: Measurment_noise}
	 Y(t_i) = y(\theta, t_i) + \epsilon(t_i)
\end{equation}}
where $y$ is the output model, which is a function of $\theta$ and time $t_i$.

Because of the difficulty of obtaining analytical solutions to the deterministic process model, experiments have been arranged whereby the vector of derivatives $d Y(t_i)/dt$ is measured rather than $Y(t_i)$ itself. In such cases, it is assumed that the unobservable error is added to the deterministic derivative $dy(\theta, t_i)/dt$ as follows:

{\footnotesize
\begin{equation}
	\cfrac{d Y(t_i)}{dt} = \cfrac{dy(\theta, t_i)}{dt} + \epsilon(t_i)
\end{equation} }

\subsection{Least squares estimation}
A least squares parameter estimation does not require prior knowledge of the distribution of unobservable errors, yield  unbiased estimates, \sout{that is, $ \mathbb{E}\{\hat{Y}(t) = y(t) \} $,} and results in the minimum variance among all linear unbiased estimators. If the observations $ Y$ for the model responses are continuous functions of time from $t = 0$ to $t = t_f$, the Markov (or "rigorous least squares") criterion is to minimize:

{\footnotesize
\begin{align}
	\phi &= \cfrac{1}{2} \int_{0}^{t_f}  \left[ Y(t) - y(\theta,t) \right]^T \Gamma^{-1} \left[ Y(t) - y(\theta,t) \right] dt \\
	\hat{\theta} &= \arg \min_{\theta \in \Theta} \phi
\end{align} }
where $\Gamma$ is the covariance matrix (or perhaps a matrix of appropriate weights), $\hat{\theta}$ is the parameter estimate, $\Theta$ is the parameter space, and $\phi$ is the time integrated value of the error squared ("integral squared error"). 
If the observations are made at discrete instants of time, $t_i$, $i = 1,2,...,n$, the Markov criterion is to minimize:

{\footnotesize
\begin{align}
	\phi &= \cfrac{1}{2} \sum_{i=1}^{n} \left[ Y(t_i) - y(\theta,t_i) \right]^T \Gamma^{-1} \left[ Y(t_i) - y(\theta,t_i) \right] \\
	\hat{\theta} &= \arg \min_{\theta \in \Theta} \phi
\end{align} }

If $\Gamma$ is a diagonal matrix, $\phi$  becomes a "weighted least squares" criterion; if $\Gamma = \sigma^2 \textbf{I}$ where $\sigma$ is a standard deviation., $\phi$ is the "ordinary least squares" criterion.

@@

{\footnotesize
\begin{flalign*}
	&{\normalsize \text{Model A - independent error:} } &&\nonumber \\
	Y(t_i) &= y(\theta, t_i) + \epsilon_i  \qquad i = 1,2,...,n&& \\
	&{\normalsize \text{Model B - cumulative data:}} && \nonumber \\
	Y(t_j) - Y(t_{j-1}) &= y(\theta, t_j) - y(\theta, t_{j-1}) + \epsilon_j, \quad j = 1,2,...,n&& \\
	&{\normalsize \text{or}} && \nonumber \\
	Y(t_i) = \sum_{j=1}^{i} &\left( Y(t_j) - Y(t_{j-1}) \right) = \sum_{j=1}^{i} \left( y(\theta, t_j) - y(\theta, t_{j-1}) \right) + \sum_{j=1}^{i} \epsilon_j &&
\end{flalign*} }

in which $\epsilon_i$ are independent random variables with $\mathbb{E}\{\epsilon_i\}$ and Var$\{\epsilon_i\} = \sigma_1^2$, a constant. The $\epsilon_j$ are also random independent variables because they represent differences such as $\epsilon_3' - \epsilon_2' = \left( \epsilon_1 + \epsilon_2 + \epsilon_3 \right) - \left( \epsilon_1 + \epsilon_2 \right) = \epsilon_3$. We assume that $\mathbb{E}\{\epsilon_j\}$ and Var$\{\epsilon_i\} = \sigma_1^2\left(y_j - y_{j-1}\right)$; i.e., the variance of $\epsilon_j$ can be a function of the test interval. The experimental point tend to stay on one side of the regression line of the best fit for Model B.

The \sout{best} unbiased estimate of the Model B is obtained by minimizing the weighted sum of squares:

{\footnotesize
\begin{equation*}
	\hat{\theta} = \arg \min_{\theta \in \Theta} \sum_{j=1}^{n} \cfrac{ \left[ \left( Y(t_j) - Y(t_{j-1}) \right) - \left( y(\theta, t_j) - y(\theta, t_{j-1}) \right) \right]^2  }{y(\theta, t_j) - y(\theta, t_{j-1})}
\end{equation*} }

\subsection{Maximum likelihood estimation}

Maximum likelihood estimation (MLE) is a method of estimating the parameters of an assumed probability distribution, given some observed data. This is achieved by maximizing a likelihood function so that, under the assumed statistical model, the observed data is most probable. The MLE has the desirable characteristics of asymptotic efficiency and normality. Each time it has been associated with the (joint) normal distribution because of mathematical convenience. Consider the joint probability density function (the likelihood function) $p(\theta |y)$ for random sample $y = y(t_1),y(t_2),...,y(t_n)$, parameter $\theta$. If a maximum of this function over all choices of $\theta$ can be found, the estimates so obtained are maximum likelihood estimates. The conditions at the maximum can be evolved incorporating prior information as follows.

The posterior probability density $p(\theta|y)$ can be expressed as the ratio of two probability densities if we make use of the analogue for continuous variables of Equation \ref{EQ: Probabiltiy_independet_b}:

{\footnotesize
	\begin{align} \label{EQ: Posterior_distribution}
		p\left( \theta| y(t_n),...,y(t_1) \right) = \cfrac{p\left( \theta, y(t_n),...,y(t_1) \right)}{p\left( y(t_n),...,y(t_1) \right)}
\end{align} }

{\footnotesize
	\begin{equation}
		\begin{split}
			p\left( \theta,y(t_n),...,y(t_1) \right) &= p\left( y(t_n)|\theta,y(t_{n-1}),...,y(t_1) \right) \\ &\cdot p\left( \theta,y(t_{n-1}),...,y(t_1) \right)
		\end{split}
\end{equation} }

{\footnotesize
	\begin{equation}
		p\left( \theta, y(t_n),...,y(t_1) \right) = p\left( \theta \right) \prod_{i=1}^{n} p\left( y(t_i)|\theta, y(t_{i-1}),...,y(t_1) \right)
\end{equation} }

Examination of Equation \ref{EQ: Measurment_noise} shows that $ Y(t_i)$ depends only on $t_i$, $\theta$ and $\epsilon(t_i)$ and is not conditioned by any previous measurement. Consequently, we can write

{\footnotesize
	\begin{equation}
		p\left( y(t_i)|\theta, y(t_{i-1}),...,y(t_1) \right) = p\left( y(t_i)|\theta \right)
\end{equation} }

{\footnotesize
	\begin{equation}
		p\left( \theta| y(t_n),...,y(t_1) \right) = \cfrac{p\left( \theta \right) \prod_{i=1}^{n} p\left( y(t_i)|\theta \right) }{p\left( y(t_n),...,y(t_1) \right)}
\end{equation} }

We can get rid of the evidence term $p\left( y(t_n),...,y(t_1) \right)$ because it's constant with respect to the maximization. Moreover, if we are lacking a prior distribution over the quantity we want to estimate, then $p(\theta)$ can be omitted. In such a case:

{\footnotesize
	\begin{equation}
		p\left( \theta| y(t_n),...,y(t_1) \right) = \prod_{i=1}^{n} p\left( y(t_i)|\theta \right) = \prod_{i=1}^{n} L\left( \theta|y(t_i) \right) 
\end{equation} }

By collecting a random sample of values of the random variable $y$ and selecting the values of $\theta$ that maximize the likelihood function $L\left( \theta | y \right)$, a function described in chapter \ref{CH: Bayes} in connection with Bayes' theorem. Such estimators, $\hat{\theta}$, are known as maximum likelihood estimators. In effect, the methods selects those values of $\theta$ that are at least as likely to generate the observed sample as any other set of values of the parameters if the probability density of the random variable $y$ were to be extensively simulated through use of the probability density $p\left( y|\theta \right)$. In making a maximum likelihood estimate, we assume that the form of the probability density (\sout{only the $\theta$ need be determined}) and that all possible values of $\theta$ are equally likely before experimentation.

%The likelihood function for the parameters given one observation is just the probability density in which the observation is regarded as a fixed number and the parameters as the variables:
%
%{\footnotesize
%	\begin{equation}
%		p\left( \theta| y(t_i) \right) = L\left( \theta|y_{t_i} \right) = p\left( y(t_i)| \theta \right)
%\end{equation} }
%where the lower case $y$'s and the number subscripts designate the value of the respective observation that is inserted into the probability density function. 
The likelihood function for the parameters based on several observations is the product of the individual functions if the observations are independent.

{\footnotesize 
	\begin{multline}
		L\left( \theta|y(t_n),...,y(t_1) \right) = \prod_{i=1}^{n} L\left( \theta|y(t_i) \right) \\
		= p\left( y(t_1)| \theta \right)p\left( y(t_2)| \theta \right) ... p\left( y(t_n)| \theta \right)
\end{multline} }

In choosing as estimates of $\theta$ the values that maximize $L$ for the given values $\left( y(t_i) \right)$, it turns out that it is more convenient to work with the $\ln L$ than with $L$ itself:

{\footnotesize 
	\begin{equation}
		\ln L = \ln p \left( y(t_1)| \theta \right) + \ln p\left( y(t_2)| \theta \right) + ... + \ln p \left( y(t_n)| \theta \right) = \sum_{i=1}^{n} \ln p\left( y(t_i); \theta \right)
	\end{equation} }

By assuming that the conditional distribution of $\bar{Y}_i$, given $y_i$, is normal, then we form the likelihood function based on the probability density:

{\footnotesize
	\begin{align} \label{EQ: MLE_Norm}
		p\left(  \theta, \sigma | y(t_n),...,y(t_1) \right) &= \prod_{i=1}^{n} \cfrac{1}{\sqrt{2 \pi} \sigma } \exp \left[ \cfrac{1}{2  \sigma } \left( Y(t_i) - y(\theta , t_i) \right)^2 \right] \nonumber \\
		L\left( \theta, \sigma | y(t_n),...,y(t_1) \right) &= \prod_{i=1}^{n} \cfrac{1}{\sqrt{2 \pi} \sigma } \exp \left[ \cfrac{1}{2  \sigma } \left( Y(t_i) - y(\theta , t_i) \right)^2 \right]
\end{align}}
where $\sigma$ is the variance

%In Equation \ref{EQ: MLE_Norm} the parameters are the variables, and the values of $Y$ and $x$ are given. Then
By taking the natural logarithm of the Equation \ref{EQ: MLE_Norm}, the final form of the objective function can be obtained:

{\footnotesize
	\begin{equation}
		\ln L = -n \ln \sqrt{2 \pi} - \cfrac{n}{2} \ln \sigma
		- \cfrac{ \sum_{i=1}^{n} \left[  Y(t_i) - y(\theta , t_i) \right]^2 }{2 \sigma^2}
	\end{equation}
}

The parameter estimation problem can be formulated as follow:

{\footnotesize
	\begin{equation}
		\begin{aligned}
			&\hat{\theta}_{MLE} &= \arg \max_{\sigma, \theta \in \Theta} \ln L = \arg \max_{\sigma,\theta \in \Theta} p(\theta|y) \\
			&\text{subject to}
			& \dot{x} = f(t,x,\theta) \\
			&& \dot{\theta} = 0 \\
			&& y = g(x)
		\end{aligned}
\end{equation} }

{\color{red}Based on the first order optimality condition,} the $\ln L$ can be maximized with respect to the vector $\theta$ by equating to zero the partial derivatives of $\ln L$ with respect to each of the parameters:

{\footnotesize
	\begin{align}\label{EQ: MLE_Basic_Solution}
		\cfrac{\partial \ln L}{\partial \theta} &= \cfrac{\partial \sum_{i=1}^{n} \ln p\left( y(t_i) | \theta \right) }{\partial \theta} = 0 
\end{align} }

Solution of Equations \ref{EQ: MLE_Basic_Solution} yield the desired estimates $\hat{\theta}$. For some models, these equations can be explicitly solved for $\hat{\theta}$ but in general no closed-form solution to the maximization problem is known or available, and an MLE can only be found via numerical optimization.

\sout{By carrying out this operation, under fairly unrestrictive conditions,} 
It can be shown that as $n$ approaches infinity the maximum likelihood estimates have the desirable asymptotic properties:

\begin{enumerate}[label=(\arabic*)]
	\item {\footnotesize $\lim\limits_{n \rightarrow \infty} \mathbb{E}\{\hat{\theta}\} = \theta$ }
	\item {\footnotesize $\left[ \sqrt{n}\left( \hat{\theta} - \theta \right) \right]$ } is normally distributed
\end{enumerate}
and for the case of two parameters:

\begin{enumerate}[resume,label=(\arabic*)]
	\item {\footnotesize $\lim\limits_{n \rightarrow \infty} \text{Var}\{\hat{\theta}\} = \cfrac{1}{n} \left[ \mathbb{E} \left\{ \left( \cfrac{\partial \ln p}{\partial \theta} \right)^2 \right\} \right]^{-1} \cfrac{1}{1-\rho^2_{\hat{\theta}}} $ } \label{Assumption_MLE}
\end{enumerate}

where $\rho^2_{\hat{\theta}}$ is the matrix of correlation coefficients.

Maximum likelihood estimates are not necessarily unbiased. Maximum likelihood estimates, however are efficient and, hence, consistent estimates. Furthermore, where a sufficient estimate can be obtained, the maximum likelihood method will obtain it. Finally, if $\hat{\theta}$ is a maximum likelihood estimator of $\theta$, the $f\left( \hat{\theta} \right)$ is a maximum likelihood estimator of $f\left( \theta \right)$, a function of $\theta$.

@@

%Under the assumption of normality, the lack of prior knowledge about the parameters $\theta^0$ and known initial conditions $y_0$. Intuitively, this selects the parameter values that make the observed data most probable. The specific value $\hat{\theta}$ that maximizes the likelihood function $L$ is called the maximum likelihood estimate.

As already discussed, the prior information about the estimated parameters can be incorporated in to the maximum likelihood estimator to form a maximum a posteriori probability (MAP) estimate. MAP can be used to obtain a point estimate of an unobserved quantity on the basis of empirical data. MAP estimation can therefore be seen as a regularization of maximum likelihood estimation. The relationship between MAP and MLE can be obtained from the Bayes' rule:

{\footnotesize
\begin{equation} 
	\begin{split} \label{EQ: MAP_probability}
		{\bf \hat{\theta}_{\text{MAP}}} &= \arg\max_{\sigma, \theta \in \Theta} p(\theta | y) \\
		&= \arg\max_{\sigma, \theta \in \Theta} \frac{p(y | \theta) p(\theta)}{p(y)} \\
		&= \arg\max_{\sigma, \theta \in \Theta} p(y | \theta) p(\theta) \\
		&= \arg\max_{\sigma, \theta \in \Theta} \ln(p(y | \theta) p(\theta)) \\
		&= \arg\max_{\sigma, \theta \in \Theta} \ln p(y | \theta) + \ln p(\theta)
	\end{split}
\end{equation} }

We assume the normal distribution of random samples, with identical variance and the prior distribution of $\theta$ given by $\mathcal{N}(\theta^0,\,\sigma_\theta^{2})$. The function to be maximized can be written as

{\footnotesize
\begin{equation} \label{EQ: MAP_L2}
	\begin{aligned}
	&\arg\max_{\sigma, \theta \in \Theta} \left[ \ln \prod_{i=1}^{n} \frac{1}{\sigma\sqrt{2\pi}}e^{-\frac{ \left( Y(t_i) - y\left( \theta, t_i \right) \right)^2 }{2\sigma^2}}
	+ \ln \prod_{j=0}^{p} \frac{1}{\sigma_\theta\sqrt{2\pi}}e^{-\frac{ \left( \theta - \theta^0 \right)^2}{2\sigma_\theta^2}} \right]  \\
	= &\arg\max_{\sigma, \theta \in \Theta} \left[- \sum_{i=1}^{n} {\frac{\left( Y(t_i) - y\left( \theta, t_i \right) \right)^2}{2\sigma^2}}
	- \sum_{j=0}^{p} {\frac{\left( \theta - \theta^0 \right)^2}{2\sigma_\theta^2}} \right]\\
	= &\arg\min_{\sigma, \theta \in \Theta} \frac{1}{2\sigma^2} \left[ \sum_{i=1}^{n} \left( Y(t_i) - y\left( \theta, t_i \right) \right)^2
	+ \frac{\sigma^2}{\sigma_\theta^2} \sum_{j=0}^{p} \left( \theta - \theta^0 \right)^2 \right] \\
	= &\arg\min_{\sigma, \theta \in \Theta} \left[ \sum_{i=1}^{n} \left( Y(t_i) - y\left( \theta, t_i \right) \right)^2 + \lambda_{L2} \sum_{j=0}^{p} \left( \hat{\theta} - \theta^0 \right)^2 \right]
	\end{aligned}
\end{equation} }

In Equation \ref{EQ: MAP_L2} some of the constants (with respect to $\theta$) were dropped and factored to simplify the expression. We can adjust the amount of regularization we want by changing $\lambda_{L2}$. Equivalently, we can adjust how much we want to weight the priors carry on the coefficients ($\theta$). If we have a very small variance (large $\lambda_{L2}$) then the coefficients will be very close to 0; if we have a large variance (small $\lambda_{L2}$) then the coefficients will not be affected much (similar to as if we didn't have any regularization).

Let's assume that the prior distribution of parameter $\theta$ is given by the Laplace distribution. The likelihood function can defined as follow:
